---
title: "R wizardry course week 5 (skeleton), 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
proc.time()
```

# Week 5 content

5.1 for loops
5.2 cycle through values
5.3 inclass problem
5.4 if statements
5.5 inclass problems
5.6 debugging loops
5.7 while loops
5.8 Next and break
5.9 Nested looping

```{r}

rm(list=ls(all=TRUE))

setwd("/Admin/desktop/R/Datasets/")
data <- read.csv("compounds_triplicates_long.csv",header=T,stringsAsFactors = FALSE)

```

## 5.1 for loops

R doesn't care if you write something 1000 times or have it copy it 1000 times
But the second is a lot easier for you

```{r}


```

for loops have two stages
1. Set a variable to equal the next value in a sequence
2. Run a set of code with that variable at that value
3. Repeat

```{r}

```

R either takes the rest of the line, the next line, or everything within { } as the looped code

```{r}

```

We can use the variable simply as a tracker to repeat code a given number of times
For example, we can replicate the rep() function

```{r}

```


## 5.2 cycle through values

The most useful thing to do with a for loop is to cycle through values

```{r}

```

## 5.3 inclass problems
Replicate the tapply() function (you don't need it to have the same formatting)

```{r}

tapply(data$mean_methane, data$compound, mean, na.rm=TRUE)

compound <- data.frame("compound" = unique(data$compound),"mean_methane" = 0)
compound


```


## 5.4 if statements

The big advantage of programming is you can have R make decisions
Allows code to be flexible, dont have to rewrite everything if you change the data
R makes binary decisions: "if data meets a condition, do this thing or else do this other thing"
if(conditional argument that leads to one F or T outcome)

```{r}

```

R either takes the rest of the line, the next line, or everything within { } as the conditional action

```{r}

```

if() is useful if we have R make the decision for us

```{r}

```

We can use the else statement for efficiency if we want an alternative thing to be done only when the first thing is not done else goes after the if() task and runs using the same rules as an if() task

```{r}

```

For simple tasks, ifelse() puts it all in one function
Need to use <- instead of =

```{r}

#Can also use ifelse() for staggered data selection


```


## 5.5 inclass problems
Use t-test to look for methane production diffrences between the salinity levels of every 10th day (starting at day 2), excluding the "BRACKISH" salinity.

```{r}

#subset dataset for every 10th day and exclude brackish


```

Now use an if statement so you don't have to subset your data by day

```{r}

```

## 5.6 debugging loops

```{r}
#Sy.sleep can help you debug your for loops by slowing it down (if combined with a print or plot function)


#print can help you figure out where you loop is breaking down


#You can also set the variable you are looping with to different value to see what happens

```

## 5.7 while loops

```{r}


#Only checks if the statement is true on each loop


#you can press escape to stop the code


```

## 5.8 Next and break

```{r}

#or


#Skip a value (iteration) using next


#Skipping nonexistant data


```

## 5.9 Nested looping

```{r}


#mixing and matching flow control


```