---
title: "R wizardry course week 7, 2017"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
rm(list=ls(all=T))
```


# Week 7 content. ggplot2 basics.


7.1 Grammar of graphics
7.2 Scatter plots
7.3 Line plots
7.4 Bar plots


## 7.1 Grammar of graphics

### What Is Grammar Of Graphics?

"A	grammar	of	graphics	is	a	tool	that	enables	us	to concisely	describe	the	components	of	a	graphic" -Hadley	Wickham.
                                
### ggplot2 and the grammar of graphics philosophy behind it 
Originally	described	in	the	work	of	Leland	Wilkinson 1 a	grammar	of graphics	gives	us	the	formal	elements	of	a	graphic	so	we	can	describe	and construct	any	statistical	graphic	and	move	beyond	named	graphics	(i.e., "scatterplot").


"The basic idea: independently specify plot building blocks and combine them to create just about any kind of graphical display you want". Those blocks are:

* **data**: must be a **data frame** for ggplot2 to work

* **aesthetic mapping (aes)**: describes the relationship (x and y, colors, fill) but doesn't supply data.

* **geometric object (geom)**: Shapes (e.g. circles and diamonds) that represent observations

* **statistical transformations (stats)**: Some geoms are indeed statistics, e.g. geom_histogram().

* **scales**: Linear scaling of x and y axes. Scales discrete and continuous.

* **coordinate system**: e.g. geom_smooth(), stat_summary()

* **position adjustments**:

* **faceting**: make multipanel plots of data subsets. Helpful to "declutter" busy graphics

"Every geom has a default statistic, every statistic has a default geom (can be overridden)" (http://ggplot2.org/resources/2007-vanderbilt.pdf).

Based on these concepts, ggplot2 makes fairly easy to produce publication quality plots with only a few lines of code.


### ggplot requirements/recommendations

* Data must be in a data frame

* Long versus wide format: Each column should be a variable and each row an observation


WARNING: 
Enhancing plot's aesthetics can be VERY ADDICTIVE!
Before start coding, set your mind on what type of plot you want (line, plot,both, what to plot, etc.). Plotting can become a time consuming task, often unnecessarily.

### Base R plots vs. ggplot2

## Histograms

Histograms in base R
```{r}

head(iris)
str(iris)
unique(iris$Species)

hist(iris$Petal.Length)
```

Histograms in ggplot
```{r}
library(ggplot2)
my_histogram <- ggplot(iris, aes(x = Petal.Length)) +
  geom_histogram()

print(my_histogram) 

```


## Boxplots

Base R

```{r}
boxplot(Petal.Length ~ Species, data = iris)
?boxplot
```

ggplot

ggplot automatically optimizes scales for improved aesthetics by deafult, ultimatelly a graphic that is easier to interpret.
```{r}
my_boxplot <- ggplot(iris, aes(x = Species, y = Petal.Length)) + 
  geom_boxplot()

print(my_boxplot)
```


## Scatter plot

```{r}
plot(Petal.Length ~ Sepal.Length, data = iris, col = Species)

```

ggplot
```{r}
my_scatter <- ggplot(data = iris, aes(x= Sepal.Length, y = Petal.Length)) +
  geom_point(aes(colour = Species))

print(my_scatter)
```


## A closer look:

Histogram:
* Geom/Stat: Histogram
* Scale: linear
* Coordinate system: cartesian

Bar plot:
* Geom: bar
* Stat: bin
* Scale: linear
* Coordinate system: cartesian

Scatter plot:
* Geom: point
* Stat: identity
* Scale: linear
* Coordinate system: cartesian

```{r, fig.height = 6, fig.width = 12, fig.align = "left"}
library(cowplot) # cowplot has many other features, not just mulitple plots in a page
plot_grid(my_histogram, my_boxplot, my_scatter, nrow = 1, ncol = 3)
detach("package:cowplot", unload = TRUE)

# ALternatively,

library(gridExtra)
grid.arrange(my_histogram, my_boxplot, my_scatter, nrow = 1, ncol = 3)

```


## Facet grids

Facets are handy to avoid busy, hard to interpret plots.


```{r}

# Horizontal facets
my_scatter + facet_wrap(~ Species)


# Horizontal facets (default) 
my_scatter + facet_wrap(~ Species, strip.position = "right")
my_scatter + facet_wrap(~ Species, strip.position = "left")

#Vertical facets
my_scatter + facet_wrap(~ Species, strip.position = "right", dir = "v")
my_scatter + facet_wrap(~ Species, strip.position = "left", dir = "v")

```


## Stats

### Stats smooth
```{r}

# Notice the colors argument (inside aes()!)
ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species)) + 
  geom_point() + 
  stat_smooth(method="lm")

```


A common mistake is to put the color argument outside aes()
```{r}

ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width), color=Species) +
  geom_point() + 
  stat_smooth(method="lm")
```



## Scales

Scales are used to modify axes and colors. 

For example:

* `facet(aes(scale = "free"))`

```{r}
my_scatter + facet_grid(Species ~ ., scales = "free_y") # notice this is facet_grid, no facet_wrap. facet_grid() makes plots for all possible result/ecombinations in the ploted variables, including empry plots. facet_wrap() only makes plots that contain data.
```

* `scale_y_continuous()` Set name, breaks, labels, limits of y-axis


```{r}

my_scatter + 
  scale_x_continuous(breaks = c(5.5, 6.5), limits = c(3,8)) +
  scale_y_continuous(breaks = c(2.5, 3.5), limits = c(1,8))

```


* `scale_x_log10()` log transform the x-axis

```{r}

my_scatter +
  scale_x_log10() +
  scale_y_continuous(limits = c(0,8))

my_scatter +
  coord_trans(x="log2", y="log2")

```


* `scale_colour_manual()` Specify colors for geoms

```{r}
ggplot(data=iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species)) + 
  geom_point(size=3) + 
  scale_colour_manual(values=c("red","blue","yellow"))
```

* Change the order of the elements in the *legend* (deafult is alphabetical)

```{r}
ggplot(iris, aes(x=Species, y=Petal.Length, fill=Species)) +
  geom_boxplot()+
  scale_fill_discrete(breaks=c("virginica","setosa","versicolor"))

```


## Other types of geoms available in ggplot2:
```{r}
ls(pattern = '^geom_', env = as.environment('package:ggplot2'))
```


## Heatmaps
```{r}
library(ggplot2)

ggplot(airquality, aes(x=Month, y=Day)) +
  geom_tile(aes(fill = Temp), color = "white") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  ylab("Day") +
  xlab("Month")
```



## 7.2 Scatter plots


```{r}
rm(list=ls(all=TRUE))


#install.packages('Cairo') #requires install Cairo in your computer first (go to cairographics.org)
#library(dplyr)
#library(ggplot2)
#library(RColorBrewer)
search()
#detach("package:RColorBrewer")

```


## Let's use ggplot on compounds_triplicate.csv dataset.

```{r}
data <- read.csv('/home/oscar/MEGAsync/R Wizardry/Materials 2018/Datasets/compounds_stats.csv', header = TRUE, stringsAsFactors = FALSE)

head(data)
str(data)
class(data) # Object class
summary(data) # summary statistics
which(is.na(data))
```

Mapping data and first layer of aesthetics. Start by defining what x and y axes will be


## Subset the data of interest

Let's select the brackish measurements for hexane treatments.

You can achieve this via usual subsetting (base R) and assing the subsetted data to a new object

```{r}
plot_data <- data[data$salinity == 'brackish' & 
                    data$group == 'treatment' &
                  data$compound == "hexane", ]

plot_data
str(plot_data)
```


or using dplyr, which doesn't require the creation of an object if using pipes

```{r}
library(dplyr)
data %>% filter(salinity == 'brackish', 
                      group == 'treatment', 
                      compound == 'hexane') # %>% ggplot() + ...

```

```{r}
library(dplyr); library(ggplot2)

data %>% filter(salinity == 'brackish', group == 'treatment', compound == 'hexane') %>%
  ggplot(data = data, aes(x = day, y = methane_mean, shape = compound))

```

Notice we the error: "Error: Mapping should be created with `aes() or `aes_()`". It's caused because a redundance in the call of "data". Instead of calling data again, we use a period (.) to specify the data.


```{r}
point_plot <- data %>% filter(salinity == 'brackish', group == 'treatment') %>% 
  ggplot(data = ., aes(x = day, y = methane_mean)) #Doesn't work with only data, x and y. Needs a shape, color (colour) or something.

point_plot
```

Notice	the	scales	are	already	present	-	ggplot	adds	default	scales	based	on	the aesthetic and	type	of	data	(continous,	discrete,	etc.). 

Now that we have the first layer ready -data mapped-, let's start adding geoms.

Add goem_point for scatter plots, then gradients of colors and/or any desired aesthetics
```{r}
#Add arguments one at the time so students can see what they are controlling
point_plot +
  geom_point(aes(shape = compound, color = salinity, size = methane_mean))

#Add gradients:
point_plot +
  geom_point(aes(shape = compound, color = salinity, size = methane_mean)) +
  scale_colour_gradient(low = "blue")

#Error: Discrete value supplied to continuous scale.

#salinity is not a numeric variable so it can't be mapped to a gradient of color.
str(data)

#Use methane_mean instead
point_plot + 
  geom_point(aes(shape = compound, color = methane_mean, size = methane_mean)) +
  scale_colour_gradient(low = "blue", high = "red")

```

To avoid calling this object everytime during modifications, out the entire code inside print():

```{r}
point_plot2 <- point_plot + 
  geom_point(aes(shape = compound, color = methane_mean, size = methane_mean)) +
  scale_colour_gradient(low = "blue", high = "red")

point_plot2


print(point_plot2 <- point_plot + 
  geom_point(aes(shape = compound, color = methane_mean, size = methane_mean)) +
  scale_colour_gradient(low = "blue", high = "red"))
  
```


Change point size into a fixed number
```{r}
point_plot +
  geom_point(aes(shape = compound, color = methane_mean), size = 5) #changing size to a number instead of a variable will set the same shape size to all plotted points.
```


Change into open symbols instead of solids
```{r}

point_plot +
  geom_point(aes(shape = compound, color = methane_mean), size = 5) +
  scale_shape_discrete(solid = TRUE)

```

Change fill and stroke. "stroke" controls the width of the peripheral circle and "size" (notice it's outside "aes()") controls the size of the point or inner part.

Notice that to accomplish this you need to use shapes like 21 to 24 from the shapes chart (http://www.sthda.com/english/wiki/r-plot-pch-symbols-the-different-point-shapes-available-in-r).


```{r}

point_plot +
  geom_point(aes(shape = compound, color = methane_mean), size = 5) +
  scale_shape_discrete(solid = TRUE) + 
  geom_point(shape = 21, colour = "black", fill = "red", stroke = 1, size = 4) # Aesthteics stated doestream overwrite those stated first.


point_plot +
  geom_point(shape = 21, colour = "black", fill = "blue", stroke = 3, size = 4)

```

Add several layers of geom_point or other geoms to get intersting combinations of shapes, colors, etc. when you have many different data. However, it migth be a better idea to stop and think if we really need all that data in a single plot.

```{r}
point_plot +
  geom_point(aes(shape = compound, color = methane_mean), size = 5) + 
  geom_point(colour = "grey90", size = 1.5) +
  geom_point(aes(shape = factor(group)))

#http://docs.ggplot2.org/current/geom_point.html
```


## 7.3. Line plots

```{r}
library(gridExtra) #for arrows among many other features

data %>% filter(salinity == 'brackish', group == 'treatment') %>%
  ggplot(., aes(x = day, y = methane_mean)) +
  geom_line(aes(color = as.factor(compound)), arrow = arrow(),
            #arrow = arrow(angle = 15, ends = "both", type = "closed"), 
            na.rm = TRUE) +
  geom_point(aes(shape = compound))


```


Let's play with some more complex data
```{r}
data %>% 
  #filter(salinity == 'brackish', group == 'treatment') %>%
  ggplot(., aes(x = day, y = methane_mean)) +
  geom_line(aes(color = as.factor(compound)), arrow = arrow(),
            #arrow = arrow(angle = 15, ends = "both", type = "closed"), 
            na.rm = TRUE) +
  geom_point(aes(shape = compound))
```


There's a problem when trying to link the points. We need to create a unique identifier for each compound_salinity_group combination:

```{r}
#Need to create a unique identifier. Add a column to the dataset that can be passed to lines for mapping

data$unique_identifier <- paste(data$compound, data$salinity, data$group, sep = " ")
head(data)

data %>% 
  #filter(salinity == 'brackish', group == 'treatment') %>%
  ggplot(., aes(x = day, y = methane_mean)) +
  geom_line(aes(color = as.factor(unique_identifier)), arrow = arrow(),
            #arrow = arrow(angle = 15, ends = "both", type = "closed"), 
            na.rm = TRUE) +
  geom_point(aes(shape = compound))
```

Another alternative to the creation of unique identifiers is to use facets
```{r}
ggplot(data = data, aes(x = day, y = methane_mean)) +
  geom_point(aes(shape = salinity), size = 2, na.rm = TRUE) + 
  geom_line(aes(color = as.factor(unique_identifier)), na.rm = TRUE) +
  #facet_wrap( ~  compound) + #, scales = 'free')
  facet_wrap( ~  compound + group, ncol = "salinity", nrow = 3) + #, scales = 'free') 
  theme(strip.text = element_text(size=12))  #controls facet title

```

At this point, it's clear we don't need all the data we're plotting. Media Only and Sterile controls, for example, have nothing to plot. Filter the data to omit these controls

Let's only keep treatment and unamended levels of "group"
```{r}
line_plot <- data %>% filter(group == c("treatment", "unamended")) %>%
  ggplot(aes(x = day, y = methane_mean))


line_plot2<- line_plot +
  geom_point(aes(shape = paste(compound, salinity)), size = 3, na.rm = TRUE) + 
  geom_line(aes(group = unique_identifier), na.rm = TRUE) +
  #facet_wrap( ~  compound) + #, scales = 'free')
  facet_wrap( ~  group) + #, scales = 'free') 
  theme(strip.text = element_text(size=12)) + #controls facet title
  #ggtitle("Methanogenci degradation of polycyclic\n aromatic compounds")
  ggtitle("Methanogenci degradation of polycyclic\n aromatic compounds") +
  labs(x = "Time (Days)", y= "Methane formation (umoles)") +
  theme(plot.title = element_text(color="black", #family = "Arial",  
                                      face="bold", size=12, hjust=0.5, vjust = 1)) +
      
      theme(axis.title.x = element_text(color="black", #family = "Arial",
                                        face="bold",
                                        size=10, vjust = -0.5, hjust=0.5,
                                        margin=margin(20,20,0,0)), 
            axis.title.y = element_text(color="black", #family = "Arial", 
                                        face="bold", 
                                        size=10, vjust = 1, hjust=0.5,
                                        margin=margin(0,20,0,0)),
            axis.text.x = element_text (colour='black', #family = 'Arial'
                                        size = 10),
            axis.text.y = element_text (colour='black', #family = 'Arial',
                                        size = 10)) +
      theme(legend.justification=c(0.3,0), 
            #legend.position=c(0.65, 0.9),
            legend.position = "bottom",
            legend.title = element_text(size = 10), 
            legend.text = element_text(colour="black", size = 10),
            legend.key.size = unit(2, "lines"),
            legend.direction = 'horizontal')

line_plot2
```

Notice how quickly the code becomes rather complex. Comment everything! It's free!

```{r}
library(RColorBrewer)

line_plot2 +
theme(panel.background = element_rect(colour = "pink")) +
#theme_bw() +
#geom_abline(intercept = 0, slope = -90)
#geom_text(data, aes(4.5, 30, label = "Pearson-R = -0.87"))
  theme(plot.margin=unit(c(0.8, 1.5, 5, 1.5), "cm")) +
  theme(#panel.grid.major = element_blank(), #element_line(color = 'black', size = 0.4),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(), #panel.border = element_rect(color = 'black', size = 0.5),
        #panel.background = element_blank(),
        axis.line.x = element_line(color = 'black'), #element_line(),
        axis.line.y = element_line(color = 'black'))

```


## 7.4 Barplots

Barplots are used with count data. Let's use mtcars dataset. See http://docs.ggplot2.org/0.9.3.1/geom_bar.html for more examples of bar_plot uses.

```{r}

diamonds

ggplot(diamonds, aes(clarity, fill=cut)) + 
  #geom_bar(position = "dodge") +
  geom_bar(position = "stack") +
  #coord_flip() +
  facet_wrap(~ cut)
  

```


## Other common types of plots

```{r}

# horizontal lines
ggplot(data = iris, aes(x= Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point() +
  geom_hline(yintercept = 5)

# Regression lines
ggplot(data = iris, aes(x= Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point() +
  geom_abline() # regression lines

# violin plots (similar to box plots but also show the probability density)
ggplot(data = iris, aes(x= Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point() +
  geom_violin(alpha=0.1) 

# Vertical lines
ggplot(data = iris, aes(x= Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point() +
geom_vline(xintercept = 5)

# Add line segments instead of a full line (use with package "grid" to form arrows)
df <- data.frame(x1 = 5, x2 = 8, y1 =4, y2 = 6)

ggplot(data = iris, aes(x= Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point() +
  geom_curve(aes(x = x1, y = y1, xend = x2, yend = y2, colour = "curve"), data = df) +
  geom_segment(aes(x = x1, y = y1, xend = x2, yend = y2, colour = "segment"), data = df) # whatever the curve and segment mean to you. Notice they're included in the legend.


# Add lines
ggplot(data = iris, aes(x= Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point() +
  geom_line()

# Add elipses (stat)
my_plot <- ggplot(data = iris, aes(x= Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point() +
  stat_ellipse(aes(color = Species))

```


## Save your plot with ggsave
```{r}

ggsave(filename = "my_plot.pdf", device = "pdf",
       plot=my_plot, height=20, width=24, units="cm")

getwd()
warnings()
```


### Challenge

Make your own scatter plot using the "economics" dataset. Use pop as your reponse variable. Apply the themes used for the bar graph from above.

```{r, eval=FALSE, include=FALSE, echo=FALSE}
head(economics);str(economics)

ggplot(data = economics, aes(x = date, y = pop))+
  geom_line(color = "#00AFBB", size = 1) # adding themes

```



```{r}
sessionInfo()
citation("ggplot2");citation("cowplot"); citation("extrafont")

```



## Tips/take home message:

Plots are meant to help with data interpretation. Busy plots achieve the opposite. Plots are meant to help to tell a story, no to make it confusing. 

If you're running out of colors in a plot, take this as a warning that you are adding too much information to a single plot. Facets are your friends!

Avoid using 3D plots (specially if you have 2D data!) and pie charts. They are not accurate and not informative enough.

## Recommedned reads:
R naming conventions: https://google.github.io/styleguide/Rguide.xml

Recommedned addition packages:
RColorbrewer (colors for plots, including color-blind friendly palettes)
Cowplot (linear and non-linear time series)
scales (easy manipulatiion of scales)

## Sources: 
* https://rpubs.com/mccannecology/53464
* http://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html
* https://learnr.wordpress.com/2010/01/26/ggplot2-quick-heatmap-plotting/
* https://www.r-bloggers.com/how-to-create-a-fast-and-easy-heatmap-with-ggplot2/